---
title: "TRIM2_3Samples"
author: "Rashi"
date: "2024-12-17"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Librarry

```{r}
library(DESeq2)
library(pheatmap)
library(dplyr)
library(RColorBrewer)
library(ggplot2)
library(ggrepel)
library(sva)
library(apeglm)
library(EnhancedVolcano)
library("calibrate")
```

# Load Count Data

You can also embed plots, for example:

```{r, echo=TRUE}
count_data <- read.csv('transcript_count_matrix.csv', header=TRUE,row.names=1)
colnames(count_data)
head(count_data)
```
# Load sample information

```{r}

sample_info <- read.csv('pheno_data.csv',header=TRUE,row.names=1)
colnames(sample_info)
head(sample_info)
```

# Making sure the row names in sample_info matches to column names in count_data
# Making sure they are in the same order?
```{r}
all(colnames(count_data) %in% rownames(sample_info))		##should be TRUE
all(colnames(count_data) == rownames(sample_info))			##should be TRUE
```

# Set factor levels
```{r}
sample_info$state <- factor(sample_info$state, levels = c("Control", "Treat"))
sample_info$batch <- factor(sample_info$batch, levels = c("1", "2", "3"))
```

# remove batch effect
```{r}
batch <- sample_info$batch
state<- sample_info$state
data <- as.matrix(count_data)
adjusted_data <- ComBat_seq(data, batch=batch, group=state)
```
#Create a deseq object and import the count data and sample information
#dds <- DESeqDataSetFromMatrix(countData = count_data, colData = sample_info, design = ~population + state)
```{r, warning=FALSE}
dds <- DESeqDataSetFromMatrix(countData = adjusted_data, 
                              colData = sample_info, 
                              design = ~state)
```
# Set reference for the treatment  vs Control
```{r}
dds$state <- factor(dds$state, levels = c("Control","Treat"))
```

# Filter the genes
```{r}
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
head(dds)
```
# Perform statical test(s) to identify differential expressed genes
```{r, echo=FALSE, warning=FALSE}
dds <- DESeq(dds)
deseq_result <- results(dds) 
deseq_result
#write.csv(deseq_result, "1_deseq_results.csv", row.names = TRUE)
```

# Get the summary
```{r, warning=FALSE}
summary(deseq_result)
```

## Change the adjPval
```{r, warning=FALSE}
deseq_result0.05 <- results(dds, alpha = 0.05) 
summary(deseq_result0.05)
```

# Change DESEq object to R object (dataframe)
```{r}
deseq_result <- as.data.frame(deseq_result0.05)
class(deseq_result)
head(deseq_result)
```

# Order the result table by increasing p value
```{r}
deseq_result_ordered <- deseq_result[order(deseq_result$padj),]
head(deseq_result_ordered)
#write.csv(deseq_result_ordered, "2_deseq_results_ordered_padj.csv", row.names = TRUE)
```

# Extract the most DEG due to the treatment.
# Select the genes with significant change in gene expression (padj<0.05 nas lg2fc <1 and > 1)
```{r}
filtered <- deseq_result %>% filter(deseq_result$padj < 0.05)
filtered <- filtered %>% filter(abs(filtered$log2FoldChange) >1 )       # log2FC
#filtered <- filtered %>% filter(abs(filtered$log2FoldChange) >0.585 )    # log1.5FC
dim(deseq_result)
dim(filtered)
```

# Save the deseq result as filename.csv
```{r}
#write.csv(deseq_result, '3_de_result.all.csv', row.names = TRUE)
#write.csv(filtered, '4_de_result_filtered.csv', row.names = TRUE)
```

# Save the normalized counts
```{r}
normalized_counts <-counts(dds,normalized=TRUE)
head(normalized_counts)
#write.csv(normalized_counts, '5_normalized_counts.csv')
```

# Visualization
## Dispersion plot
```{r}
plotDispEsts(dds) 
```

## PCA plot
## It is a dimensionality reduction technique and in gene expression analysis, it can be used to explain the variance in
## in gene expression datasets.
## To generate a PCA plot, we will perform variance stabilization transformation for which we will use vst function in deseq and
## then we will transformed values to plot pca.
```{r}
vsd <- vst(dds,blind=FALSE, fitType='mean')
pcaData <- plotPCA(vsd, intgroup=c("state", "batch"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=state, shape=batch)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()

ggplot(pcaData, aes(PC1, PC2, color=state, shape=batch)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed() +
  theme_minimal() +  # Start with minimal theme
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    text = element_text(size = 12),  # Adjust general text size
    plot.title = element_text(size = 14, face = "bold"),  # Adjust title font
    axis.title = element_text(size = 12, face = "bold"),  # Adjust axis label font
    axis.text = element_text(size = 12)  # Adjust axis text font size
  )
```

## Heatmap
## Heatmap of sample-to-sample distance matrix (with clustering) based on normalized count.
```{r}
sampleDists <- dist(t(assay(vsd)))									### Generate distance matrix
sampleDistMatrix <- as.matrix(sampleDists)							### Generate sample distance matrix
colnames(sampleDistMatrix)
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)		### Set color Scheme
pheatmap(sampleDistMatrix, clustering_distance_rows=sampleDists, 
         clustering_distance_cols=sampleDists, col=colors)


## Heatmap of log transformed normalized count. We will use top 500 genes
#top_hits <- deseq_result[order(deseq_result$padj), ][1:100,]

top_hits <- filtered[order(filtered$padj), ][1:829,]
top_hits <- row.names(top_hits)
top_hits
rld <- rlog(dds,blind=FALSE)
annot_info <- as.data.frame(colData(dds)[,c('batch','state')])
pheatmap(assay(rld)[top_hits,],
         scale = "row",        # Scale rows (genes)
         col = colorRampPalette(c("blue", "white", "red"))(100),  # Color scheme
         breaks = seq(-2, 2, length.out = 101),
         margins = c(5, 10),   # Adjust margins for labels
         main = "Heatmap of Top Genes",
         xlab = "Sample ID",
         ylab = "Genes",
         cex.axis = 1,
         show_colnames = TRUE,
         show_rownames = FALSE,
         fontsize_row = 8,
         cluster_cols = TRUE,
         cluster_rows = TRUE, annotation_col=annot_info)
```
## Heatmap of zscores. 
```{r}
cal_z_score <- function(x) {(x-mean(x)) / sd(x)}
zscore_all <- t(apply(normalized_counts, 1, cal_z_score))
zscore_subset <- zscore_all[top_hits,]
pheatmap(zscore_subset)
```

## Count Plot
```{r}
plotCounts(dds, gene=which.min(deseq_result$padj), intgroup="state")

```
## Boxplot
```{r}
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2)
```


## MA Plot
```{r}
plotMA(dds,ylim=c(-2,2))
```

## remove the noise
```{r, warning=FALSE}
resLFC <- lfcShrink(dds,coef="state_Treat_vs_Control", type="normal")
plotMA(resLFC,ylim=c(-2,2))


res_df <- as.data.frame(resLFC)
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(color = ifelse(res_df$padj < 0.05, "red", "black")) +
  labs(x = "log2 Fold Change", y = "-log10(adjusted p-value)", title = "Volcano Plot") +
  theme_minimal()
```

## Enhanced Volcano plot
```{r}
gene_names <- sapply(strsplit(rownames(deseq_result), "\\|"), function(x) x[2])
EnhancedVolcano(deseq_result,
                lab = gene_names,
                x = 'log2FoldChange',
                y = 'pvalue',
                title = 'transcript_TRIM2',
                pCutoff = 0.05,
                FCcutoff = 2,
                pointSize = 1.0,
                labSize = 3.0,  # Increase label size
                titleLabSize = 18,  # Increase title label size
                axisLabSize = 14,  # Increase axis label size
                xlim=c(-15,25), 
                ylim=c(0.0,30),
                legendLabSize = 12)  # Increase legend label size
```


##Expression of TRIM family gene
```{r}
new_column_name <- "ID"
#deseq_result[, new_column_name] <- rownames(deseq_result)
filtered[, new_column_name] <- rownames(filtered)

# Filter rows containing the word "trim" in any column
filtered_data <- filtered[grep("trim", filtered$ID, ignore.case = TRUE), ]

top_hits <- row.names(filtered_data)
top_hits

# Extract expression levels for filtered data
rld <- rlog(dds,blind=FALSE)
filtered_expression <- assay(rld)[top_hits,]

# Extract gene names
genes <- rownames(filtered_expression)

# Iterate over each gene
for (gene in genes) {
  # Extract expression levels for the current gene
  gene_expression <- filtered_expression[gene, ]
  
  # Create a bar plot for the current gene
  barplot(gene_expression,
          main = paste(gene),
          xlab = "Samples",
          ylab = "Expression",
          col = c("purple", "yellow"), # Different colors for control and treatment
          ylim = c(0, max(filtered_expression) * 1.2),
          beside = TRUE, # Plot bars beside each other
          names.arg = c("G1_C1", "G1_T1", "G2_C2", "G2_T2", "G3_C3", "G3_T3")) # Sample names
  
  # Add legend
  legend("topright", legend = c("Control", "Treatment"), fill = c("purple", "yellow"))
}
```

#### GO Analysis
#RUN DAIVD with your genes and save data sheet as txt and import dataset from txt into R as data frame
```{r}
bp_trim2 <- read.delim("5_bp_trim2_gene.txt")
filtered_data <- subset(bp_trim2, FDR < 0.05)

library(ggplot2)
library(dplyr)

# Sort the filtered data based on Fold.Enrichment
sorted_data <- filtered_data %>% arrange(desc(Fold.Enrichment))

# Reorder the levels of the Term factor based on the sorted Fold.Enrichment
sorted_data$Term <- factor(sorted_data$Term, levels = sorted_data$Term)

# Create the lollipop chart
# Create the lollipop chart
lollipop_plot <- ggplot(sorted_data, aes(x = Fold.Enrichment, y = Term, color = FDR, size = 1/FDR)) +
  geom_segment(aes(xend = 0, yend = Term), size = 1) +  # Draw the line segment with constant size
  geom_point(shape = 19, aes(size = 1/FDR), alpha = 0.6) +  # Draw the lollipop points with size mapped to Count
  scale_color_gradient(low = "red", high = "blue") +  # Color gradient for FDR
  labs(x = "Fold Enrichment", y = "Term", color = "FDR", size = "FDR") +  # Axis labels
  theme_minimal() +  # Minimal theme
  theme(legend.position = "bottom")  # Position of the legend

# Show the plot
print(lollipop_plot)
```

